Regional Inequalities and economic growth; cross section and times series analysis

### 2023-11-20 MSB104_GR_1_Final_Assignment_research_article.qmd
Small changes in text

---
title: "Regional Inequalities and economic growth; cross section and times series analysis"
author: "Anine Therese Karlsen & Mona Lisa Jones"
format: pdf
abstract: "This assignment aims to acquire, process, and analyze sub-national GDP and population data for a subset of European countries. Firstly (assignment 1), Calculateing GDP per capita and exploreing regional inequity using various descriptive statistics and visualizations.Secondly (assignment 2), examining the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010."
editor: visual
bibliography: [Regional_inequality_23.bib, econometrics.bib]
page-layout: full
number-sections: true
papersize: a4
csl: apa.csl
---

@@

Lastly, potential limitations may arise due to the use of ChatGPT and ChatGPT PDF, which have been utilized as tools to gain understanding and extract knowledge from academic papers, as well as to explain econometric terms and definitions. The text generated by these tools has been cross-checked to the best of our knowledge with our textbook and Andre Seidel, lecturer and author of 'Regional Inequality, Convergence, and Its Determinants -- A View from Outer Space' (2017). Despite these precautions, there may have been instances where incorrect information was provided, and some errors could have been overlooked. The text has been meticulously rewritten to fit our style of writing, ensuring consistency and accuracy throughout our study.

@@

Why some places grow and prosper is a fundamental question in social science research. This question motivated Adam Smith's work, *An Inquiry into the Nature and Causes of the Wealth of Nations,* (Smith [1863](https://link-springer-com.galanga.hvl.no/article/10.1007/s11187-014-9574-4#ref-CR32 "Smith, A. (1863). An inquiry into the nature and causes of the wealth of nations. Edinburgh: A. & C Black.")) and has been a major influence of the *new economic geography* [@feldman2014]*. Furthermore,* regional growth and inequality have been important topics of research in Europe for many years. In recent decades, the European Union has implemented policies aimed at reducing regional disparities and promoting economic growth across its member states. Furthermore, despite these efforts, significant disparities in income and economic development persist across regions in Europe [@lessmann2017].

@@

Regional growth in Europe exhibits a complex pattern, influenced significantly by factors such as geography, human capital, and economic policies. Lessmann and Seidel's research offers insights into these dynamics, highlighting the persistence of income disparities, across regions despite overarching economic advancement [@lessmann2017]. Wealthier nations like Denmark tend to experience quicker regional convergence, a phenomenon supported by Gennaioli et al.'s findings [@gennaioli2014]. However, this is not uniformly experienced across all European countries, with Eastern European nations such as Hungary and Slovakia we may expect different growth trajectories, particularly after their EU accession.

@@

EU membership provided Slovakia and Hungary with resources for economic recovery, while Western European countries, including Denmark, Portugal, and France, likely experienced significant economic declines due to the financial crisis, a pattern possibly mirrored in the recent COVID-19 pandemic. These complexities highlight the importance of a detailed analysis to understand regional economic growth and inequality in challenging times.

@@

@lessmann2017 regional inequality study address the importance of studying regional inequality. Its potential consequences, such as political tensions that can undermine social and political stability [@lessmann2017]. The paper also discusses the relationship between regional inequality and personal income inequality and conflict risk.

@@

Urbanization emerges as a crucial determinant, where bustling, populous around urban centres usually enjoy elevated income levels, contrasting starkly with the often-stark disparities witnessed in more rural and secluded areas [@lessmann2017].

@@

[@lessmann2017] discusses the determinant of Education in the context of human capital. The study mentions that human capital is the most important determinant of differences in regional development within countries. The study also notes that human capital creates positive externalities, which can spill over regional boundaries. Moreover, the quality of human capital within countries, as measured by the secondary-school enrolment rate, can decrease regional inequality by facilitating regional spillovers and convergence. Lastly, promoting internal migration.

@@

Furthermore, [@lessmann2017] provides us with insights into the inverted U-shaped relationship between regional inequality and the level of economic development in different country groups [@lessmann2017]. The inverted U-shape also referred to as Kuznets Curve suggests that as an economy develops, regional inequality initially rises, reaches a peak, and then starts to decline.

@@

@iammarino2019 discusses regional inequalities in European countries. It presents evidence that inter-regional inequality has been increasing in the European Union since the turn of the millennium. The authors argue that this is due to the existence of several groups of regional economies in Europe that are structurally very different from one another [@iammarino2019]. Additionally, the authors argue that there are some countries in the EU that are more evenly developed than others, and that a map of underperformance or over-performance means less in a high-income but evenly developed country, such as in Denmark in our case [@iammarino2019]. On the other hand, France has been a subject of discussion and policy making for many years, regarding its well-known pockets of extreme wealth in regions such as Cap d'Antibes.

@@

# Data

Trough Eurostat, we download the datasets nama_10r_3gdp , demo_r_pjanggr3, demo_r_mlifexp, tran_r_net and edat_lfse_04 as csv files and filted the data by our preferences before downloading it. Furthermore, we filtered the dataset by choosing the years 2000 to 2020. Then, we selected the NUTS requiered levels region for Portugal, France, Hungary, Slovakia and Denmark. Finally, we specified the data to be in million Euro.

@@

This data set (edat_lfse_04) contains population by educational attainment level, sex and NUTS 2 regions (%). Including data on the highest level of education successfully completed by the individuals of a given population. Furthermore, data on young people neither in employment nor in education and training -- NEET, early leavers from education and training and the labour status of young people by years since completion of highest level of education. Calculated as annual averages of quarterly EU Labour Force Survey data (EU-LFS).

@@

# Econometric Approach

@@

@@

>>>>

---
title: "Regional Inequalities and economic growth; cross section and times series analysis"
author: "Anine Therese Karlsen & Mona Lisa Jones"
format: pdf
abstract: "This assignment aims to acquire, process, and analyze sub-national GDP and population data for a subset of European countries. Firstly (assignment 1), Calculateing GDP per capita and exploreing regional inequity using various descriptive statistics and visualizations.Secondly (assignment 2), examining the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010.Lastly exploring alternative functional forms and implementing panel estimation to analzise the relationship between regional development inequality"
editor: visual
bibliography: [Regional_inequality_23.bib, econometrics.bib]
page-layout: full
number-sections: true
papersize: a4
csl: apa.csl
---

@@

Furthermore, exploring alternative functional forms and implementing panel estimation to analzise the relationship between regional development inequality (***Assignment_3***).

Lastly, it is worth noting potential limitations may arise due to the use of ChatGPT and ChatGPT PDF, which have been utilized as tools to gain understanding and extract knowledge from academic papers, as well as to explain econometric terms and definitions. The text generated by these tools has been cross-checked to the best of our knowledge with our textbook and Andre Seidel, lecturer and author of 'Regional Inequality, Convergence, and Its Determinants -- A View from Outer Space' (2017). Despite these precautions, there may have been instances where incorrect information was provided, and some errors could have been overlooked. The text has been meticulously rewritten to fit our style of writing, ensuring consistency and accuracy throughout our study.

@@

Why some places grow and prosper is a fundamental question in social science research. This question motivated Adam Smith's work, *An Inquiry into the Nature and Causes of the Wealth of Nations,* (Smith [1863](https://link-springer-com.galanga.hvl.no/article/10.1007/s11187-014-9574-4#ref-CR32 "Smith, A. (1863). An inquiry into the nature and causes of the wealth of nations. Edinburgh: A. & C Black.")) and has been a major influence of the *new economic geography* [@feldman2014]*. Furthermore,* regional growth and inequality have been important topics of research in Europe for many years. In recent decades, the European Union implemented policies aimed at reducing regional disparities and promoting economic growth across its member states. Furthermore, despite these efforts, significant disparities in income and economic development persist across regions in Europe [@lessmann2017].

@@

Regional growth in Europe exhibits a complex pattern, influenced significantly by factors such as geography, human capital, and economic policies. Lessmann and Seidel's research offers insights into these dynamics, highlighting the persistence of income disparities, across regions despite overarching economic advancement [@lessmann2017]. Wealthier nations like Denmark tend to experience quicker regional convergence, a phenomenon supported by Gennaioli et al.'s findings [@gennaioli2014].

@@

EU membership provided Slovakia and Hungary with resources for economic recovery, while Western European countries, including Denmark, Portugal, and France, likely experienced significant economic declines due to the financial crisis, a pattern possibly mirrored in the recent COVID-19 pandemic. In particular Portugal as the crisis hit them harder. These complexities highlight the importance of a detailed analysis to understand regional economic growth and inequality in challenging times.

@@

@lessmann2017 regional inequality study address the importance of studying these aspects. Its potential consequences, such as political tensions that can undermine social and political stability [@lessmann2017]. The paper also discusses the relationship between regional inequality and personal income inequality and conflict risk.

@@

Urbanization emerges as a crucial determinant, where bustling, populous around urban centers usually enjoy elevated income levels, contrasting starkly with the often-stark disparities witnessed in more rural and secluded areas [@lessmann2017].

@@

[@lessmann2017] also discusses the determinant of Education in the context of human capital. The study mentions that human capital is the most important determinant of differences in regional development within countries. The study also notes that human capital creates positive externalizes, which can spill over regional boundaries. Moreover, the quality of human capital within countries, as measured by the secondary-school enrollment rate, can decrease regional inequality by facilitating regional spillovers and convergence. Lastly, promoting internal migration.

@@

Furthermore, @lessmann2017 provides us with insights into the inverted U-shaped relationship between regional inequality and the level of economic development in different country groups [@lessmann2017]. The inverted U-shape also referred to as Kuznets Curve suggests that as an economy develops, regional inequality initially rises, reaches a peak, and then starts to decline.

@@

@iammarino2019 discusses regional inequalities in European countries. It presents evidence that inter-regional inequality has been increasing in the European Union since the turn of the millennium. The authors argue that this is due to the existence of several groups of regional economies in Europe that are structurally very different from one another [@iammarino2019]. Additionally, the authors argue that there are some countries in the EU that are more evenly developed than others, and that a map of under performance or over-performance means less in a high-income but evenly developed country, such as in Denmark in our case [@iammarino2019]. On the other hand, France has been a subject of discussion and policy making for many years, regarding its well-known pockets of extreme wealth in regions such as Cap d'Antibes.

@@

# Data and Descriptive

Trough Eurostat, we download the datasets nama_10r_3gdp , demo_r_pjanggr3, demo_r_mlifexp, tran_r_net and edat_lfse_04 as csv files. Selecting our countries of study and timeperiod. Selecting the required NUTS level and specified the data to be in million Euro.

@@

This data set (edat_lfse_04) contains population by educational attainment level, sex and NUTS 2 regions (%). Including data on the highest level of education successfully completed by the individuals of a given population. Furthermore, data on young people neither in employment nor in education and training -- NEET, early leavers from education and training and the labor status of young people by years since completion of highest level of education. Calculated as annual averages of quarterly EU Labor Force Survey data (EU-LFS).

@@

@@

# Cross sectional estimates

@@

# Alternative functional forms and panel estimation

## Part 3A Testing Development effects across subsets

## Part 3B Exploring Alternative Functional Forms

## Part 3C Hetroskedasticity Testing and causality Discussion

## Part 3D Panel Estimates
